Revealing transient strain in geodetic data with Gaussian process regression

### 2017-09-26 RevisedSubmission/hines_hetland_2017_gji.tex
did more

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. Transient displacements have an easting and northing component, $\vec{u} = (u_\mathrm{e}, u_\mathrm{n})$, and we make a prior assumption that each component of $\vec{u}$ is a Gaussian process with zero mean and a generic covariance function $\mathrm{cov}\left(u_i(p),u_i(p')\right) = C_{u_i}(p,p')$. Using a more concise notation, we describe each component of $\vec{u}$ as $u_i \sim \mathcal{N}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}  
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. We also require $u_i$ to be spatially and temporally differentiable (See \citet[sec. 2.2]{Adler1981} for a formal definition of differentiation for Gaussian processes). This requirement is satisfied if the derivatives $\partial^2 X_i(\vec{x},\vec{x}')/ \partial x_j \partial x_j'$ and $\partial^2 T_i(t,t') / \partial t \partial t'$ exist everywhere. The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe, and we discuss this matter in Section \ref{sec:SignalModel}.

@@

We combine the prior on transient displacements with GNSS displacement data to form a posterior estimate of transient displacements. We ignore any covariances between the easting and northing components, which reduces the computational complexity of our analysis because we can determine each component of the posterior independently. However, we are inherently assuming that the principle components of transient displacements and the noise processes are aligned with the cardinal directions, which is admittedly an arbitrary assumption. We then form each component of the posterior, denoted as $\hat{u}_i$, by updating $u_i$ with the fact that $\mitbf{d}^*_i$ has been drawn from the random vector $\mitbf{d}_i$. That is to say $\hat{u}_i = u_i |\mitbf{d}^*_i$. If $\kappa$ is kept finite then, from \citet[sec. 2.2]{Rasmussen2006}, we find that $\hat{u}_i$ is distributed as $\mathcal{N}(\mu_{\hat{u}_i},C_{\hat{u}_i})$, where

@@

The posterior transient displacements can be evaluated at any $p$. Furthermore, $\hat{u}_i$ is spatially and temporally differentiable, allowing us to formulate $\dot\varepsilon$, which can be evaluated at any $p$. The components of $\dot\varepsilon$ can be written as 
\begin{equation}\label{eq:StrainRate}
\dot\varepsilon_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).
\end{equation}
Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian processes $u_i$ and $u_j$, we know that $\dot\varepsilon_{ij}$ is also a Gaussian process. The mean and covariance functions for $\dot{\varepsilon}_{ij}$ are given by
\begin{equation}\label{eq:StrainMean}
\mu_{\dot\varepsilon_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)
\end{equation} 
and
\begin{equation}\label{eq:StrainCov}
C_{\dot\varepsilon_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right),
\end{equation} 
respectively.

>>>>

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. Transient displacements have an easting and northing component, $\vec{u} = (u_\mathrm{e}, u_\mathrm{n})$, and we make a prior assumption that each component of $\vec{u}$ is a Gaussian process with zero mean and a generic covariance function $\mathrm{cov}\left(u_i(p),u_i(p')\right) = C_{u_i}(p,p')$. Using a more concise notation, we describe each component of $\vec{u}$ as $u_i \sim \mathcal{N}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}  
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. We also require that the derivatives $\partial^2 X_i(\vec{x},\vec{x}')/ \partial x_j \partial x_j'$ and $\partial^2 T_i(t,t') / \partial t \partial t'$ exist everywhere, which ensures that $u_i$ is spatially and temporally differentiable (See \citet[sec. 2.2]{Adler1981} for a formal definition of differentiation for Gaussian processes). The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe.  To keep this section sufficiently general, we hold off on specifying $X_i$ and $T_i$ until Section \ref{sec:SignalModel}, where we estimate transient strain from slow slip events in the Pacific Northwest.

@@

We combine the prior on transient displacements with GNSS displacement data to form a posterior estimate of transient displacements. We ignore any covariances between the easting and northing components, which reduces the computational complexity of our analysis because we can determine each component of the posterior independently. However, we are inherently assuming that the principle components of transient displacements and the noise processes are aligned with the cardinal directions, which is admittedly an arbitrary assumption. We then form each component of the posterior, denoted as $\hat{u}_i$, by updating $u_i$ with the fact that $\mitbf{d}^*_i$ has been sampled from the random vector $\mitbf{d}_i$. That is to say $\hat{u}_i = u_i |\mitbf{d}^*_i$. If $\kappa$ is kept finite then, from \citet[sec. 2.2]{Rasmussen2006}, we find that $\hat{u}_i$ is distributed as $\mathcal{N}(\mu_{\hat{u}_i},C_{\hat{u}_i})$, where

@@

We can use eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) to evaluate $\hat{u}_i$ at any $p$. Furthermore, since $\hat{u}_i$ is spatially and temporally differentiable, we can use $\hat{u}_i$ to formulate $\dot\varepsilon$, which can also be evaluated at any $p$. The components of $\dot\varepsilon$ can be written as 
\begin{equation}\label{eq:StrainRate}
\dot\varepsilon_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).
\end{equation}
Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian processes $u_i$ and $u_j$, we know that $\dot\varepsilon_{ij}$ is also a Gaussian process. The mean and covariance functions for $\dot{\varepsilon}_{ij}$ are given by
\begin{equation}\label{eq:StrainMean}
\mu_{\dot\varepsilon_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)
\end{equation} 
and
\begin{equation}\label{eq:StrainCov}
C_{\dot\varepsilon_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right),
\end{equation} 
respectively.
